Revealing transient strain in geodetic data with Gaussian process regression

### 2017-11-25 ResponseToReviewers/ResponseToReviewers.tex

\textbf{Page 4: Stochastic model: It would seem that the assumption of no
covariance between the north and east components is not a very good
one since there is a tectonic framework in which the transients occur
and its unlikely this framework aligns along the Cardinal directions.
Maybe add a statement of impact of the neglect and possibly the idea
of reorienting the “axes” to align with the tectonic framework e.g.,
perpendicular to the subduction zone interface in this case?}

This is a good point that we neglected to mention in the previous
draft. We have added a paragraph to Section 2 that explains the
reasoning and implications for this assumption. The paragraph starts
with "It should be noted that we have ignored any covariances ...".

\textbf{Page 4/5 Equation (3): Is it worth discussing at this point whether
the estimates of u and n (eta) can be separated without explicit prior
knowledge (which we may not have).  For example, orbit modeling errors
on one satellite (e.g., due to an unknown yaw problem in the
spacecraft) wilt generate a spatially and temporally correlated error
in $d_{ij}$.  How does this (stochastically non-modeled) error not project
in the u estimates?}

At this point in the text, it is assumed that the statistical
properties of u and eta are known, although not yet specified. So I do
not think this would be an appropriate place to discuss a scenario in
which we do not know u or eta. I also think that it is clear that our
results are contingent on having a good model for u and eta. Asking
whether the strain signal can be resolved without a prior statistical
model for u or eta is asking whether the signal can be separated from
the noise if one does not know anything about the noise or the signal.
I think it goes without stating that signal can not be extracted from
observations without a model of the noise or an assumption of the
signal sought.

In Section 3.1 and 3.2, we discuss how we choose a model for u and
eta. The key piece of "prior knowledge" that allows us to discern u
from eta is that u is negligible for inland stations. This allowed us
to isolate and characterize eta first, and then we characterized u. If
we allowed eta to be spatially correlated (which we did not), then we
would have conceivably been able to describe the spatially and
temporally correlated noise from, for example, orbit modeling errors.

@@

This sentence has been removed in the revised draft

\textbf{6+56: I suspect that equation (15) took some effort to derive; could
some details go into an Appendix?}

I added a clearer derivation in the revised draft.

\textbf{Figure 6: I like that the strain uncertainties can be shown, but is
there a reason not to plot principal strains using arrows, showing
errors by an ellipse around the tips?}

I have put a lot of thought into representing strain, and I recognize
that strain is conventionally represented by showing the principal
strains axes. However, there is no good way of showing the
uncertainties for the principal strain axes. For example, consider a
perfectly dilational strain that is perturbed by a small amount of
noise. The orientation of principal strains will be highly uncertain,
but the magnitudes of the principal strains will be well constrained.
If we were to draw some sort of confidence interval around the tips of
the principal strains, then they would have a crescent shape (not an
ellipse). It is unclear how I would write a program to draw these
confidence intervals.

Instead of showing the principal strains, I am showing the normal
strains as a function of azimuth. For a given azimuth, the uncertainty
in the normal strain is computed by linear propagation of the
uncertainties in the strain. In my opinion, it is much easier to show
the uncertainties in an azimuthal representation of strain.

### 2017-11-26 revised_source_files/hines_hetland_2017_gji.tex
ready to resubmit

Crustal strain rates are fundamentally important quantities for
assessing seismic hazard. Knowing where and how quickly strain is
accumulating gives insight into where we can expect stored elastic
energy to be released seismically. Estimates of secular, or long-term,
strain rates have been used to constrain seismic hazard models such as
UCERF3 \citep{Field2014}. Transient crustal strain, which is anomalous
strain caused by geophysical processes such as slow slip events (SSEs)
or postseismic deformation, is also relevant for assessing seismic
hazard. While transient strain itself is not damaging, there is a
possibility that it can trigger major earthquakes
\citep{Roeloffs2006,Freed2001}.

@@

The function $C_{u_i}$ must be positive definite in order to be a
valid covariance function. By definition, $C_{u_i}$ is a positive
definite function if the matrix $[C_{u_i}(p,p')]_{(p,p') \in \points
\times \points}$ is positive definite for any set of points $\points$
\citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be
separated into spatial and temporal functions as
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}(p,p') = C_{u_i}\left((\pos,t),(\pos',t')\right) 
              = X_i(\pos,\pos')T_i(t,t').
\end{equation}   
As long as the functions $X_i$ and $T_i$ are positive definite,
$C_{u_i}$ is guaranteed to also be positive definite \citep[sec.
4.2.4]{Rasmussen2006}. We also require that the derivatives
$\partial^2 X_i(\pos,\pos')/ \partial x_j \partial x_j'$ and
$\partial^2 T_i(t,t') / \partial t \partial t'$ exist. This ensures
that $u_i$ is spatially and temporally differentiable, allowing us to
compute transient strain rates (See \citet[sec. 2.2]{Adler1981} or
\citet[sec. 10A]{Papoulis1991} for a definition of stochastic
differentiation and the conditions for differentiability).

@@

We provide an example to give the prior on transient displacements a
more tangible meaning. we can use a squared exponential function for
$X_i$ and $T_i$,
\begin{equation}\label{eq:ExampleCov}
X_i(\pos,\pos') = \exp\left(\frac{-||\pos - \pos'||_2^2}{2\ell^2}\right), \ \ \ \ 
T_i(t,t') = \phi^2 \exp\left(\frac{-|t - t'|^2}{2\tau^2}\right),
\end{equation}
which satisfies our requirements for positive definiteness and
differentiability. The parameters $\ell$ and $\tau$ control the
length-scale and time-scale, respectively, of realizations of $u_i$.
The parameter $\phi$, which we have arbitrarily chosen to incorporate
into $T_i$ rather than $X_i$, controls the amplitude of realizations
of $u_i$. Ideally, we want realizations of $u_i$ to have a
length-scale, time-scale, and amplitude that resemble what we expect
for the true transient displacements. In Figure \ref{fig:PriorDemo}A,
we show $C_{u_i}$ using the squared exponential function for $X_i$ and
$T_i$ and the parameters $\ell = 100$ km, $\tau = 10$ days, and $\phi
= 1$ mm. A single realization of $u_i$ corresponding to that choice of
covariance function is shown in Figure \ref{fig:PriorDemo}B (See
\citet[sec. A3]{Rasmussen2006} for details on drawing realization from
Gaussian processes). While the squared exponential function is a
commonly used covariance function for GPR, it is not appropriate for
every application. The appropriate choice for $X_i$ and $T_i$ may vary
depending on the geophysical signal we are trying to describe. To keep
this section sufficiently general, we hold off on specifying $X_i$ and
$T_i$ until Section \ref{sec:SignalModel}, where we estimate transient
strain from slow slip events in the Pacific Northwest.

@@

It should be noted that we have ignored any covariances between the
easting and northing components of $\vec{u}$ and $\vec{\data}$.
This simplification reduces the computational complexity of evaluating
the posterior transient displacements because each component can be
evaluated independently. However, we are inherently assuming that the
principle axes describing the distribution of $\vec{u}(p)$ and
$\vec{d}^{(jk)}$ are aligned with the cardinal directions, which is
admittedly an arbitrary assumption.

@@

The posterior transient displacements are spatially and temporally
continuous, and we can use eqs. (\ref{eq:PosteriorMean3}) and
(\ref{eq:PosteriorCov3}) to evaluate $\post_i$ at any $p$.
Furthermore, $\post_i$ is spatially and temporally differentiable,
allowing us to formulate $\strain$ at any $p$ that we may be
interested in. The components of $\strain$ can be written as
\begin{equation}\label{eq:StrainRate}
\strain_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} 
                  \left(\frac{\partial \post_i(p)}{\partial x_j} +  
                        \frac{\partial \post_j(p)}{\partial x_i} \right).
\end{equation}
Since eq. (\ref{eq:StrainRate}) is a linear operation on the Gaussian
processes $\post_i$ and $\post_j$, we know that $\strain_{ij}$ is also
a Gaussian process. From \citet[sec. 10]{Papoulis1991}, we find that
the mean and covariance functions for $\dot{\varepsilon}_{ij}$ are
\begin{align}\label{eq:StrainMean}
\mu_{\strain_{\e\e}}(p) &= \frac{\partial^2 \mu_{\post_\e}(p)}{\partial t \, \partial x_\e} \\
\mu_{\strain_{\n\n}}(p) &= \frac{\partial^2 \mu_{\post_\n}(p)}{\partial t \, \partial x_\n} \\
\mu_{\strain_{\e\n}}(p) &= \frac{1}{2}\frac{\partial}{\partial t}\left(
                                    \frac{\partial \mu_{\post_\e}(p)}{\partial x_\n} + 
                                    \frac{\partial \mu_{\post_\n}(p)}{\partial x_\e} \right)
\end{align} 
and
\begin{align}\label{eq:StrainCov}
C_{\strain_{\e\e}}(p,p') &= \frac{\partial^4 C_{\post_\e}(p,p')}{\partial t \, \partial t' \, \partial x_\e \, \partial x'_\e} \\
C_{\strain_{\n\n}}(p,p') &= \frac{\partial^4 C_{\post_\n}(p,p')}{\partial t \, \partial t' \, \partial x_\n \, \partial x'_\n} \\
C_{\strain_{\e\n}}(p,p') &= \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                        \frac{\partial^2 C_{\post_\e}(p,p')}{\partial x_\n \, \partial x'_\n} + 
                                        \frac{\partial^2 C_{\post_\n}(p,p')}{\partial x_\e \, \partial x'_\e} \right),
\end{align} 
respectively. The above derivatives of $\mu_{\post_i}$ and
$C_{\post_i}$ are computed analytically by replacing the terms
$C_{u_i}(p,\points)$, $C_{u_i}(\points,p')$, and $C_{u_i}(p,p')$ in
eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) with their
appropriate derivatives. For example, the mean and covariance
functions for $\strain_{\e\e}$ can be written more verbosely as
\begin{equation}\label{eq:StrainMeanEE}
\mu_{\strain_{\e\e}}(p) = \left[\begin{array}{cc}
                          \frac{\partial^2 C_{u_\e}\left( p,\points \right)}{\partial t \, \partial x_\e} & \zeros \\
                          \end{array}\right]
                          \left[\begin{array}{cc}
                          C_{u_\e}(\points,\points) + \mitbf{C}_{\eta_\e} & \G \\
                          \G^T  & \zeros \\
                          \end{array}\right]^{-1}
                          \left[\begin{array}{c}
                          \data^*_\e \\
                          \zeros \\
                          \end{array}\right]
\end{equation}    
and
\begin{multline}\label{eq:StrainCovEE}
C_{\strain_{\e\e}}(p,p') = \frac{\partial^4 C_{u_\e}(p,p')}{\partial t \, \partial t' \, \partial x_\e \, \partial x'_\e} - \\
                           \left[\begin{array}{cc}
                           \frac{\partial^2 C_{u_\e}\left( p,\points \right)}{\partial t \, \partial x_\e} & \zeros \\
                           \end{array}\right]
                           \left[\begin{array}{cc}
                           C_{u_\e}(\points,\points) + \mitbf{C}_{\eta_\e} & \G \\
                           \G^T  & \zeros \\
                           \end{array}\right]^{-1}
                           \left[\begin{array}{c}
                           \frac{\partial^2 C_{u_\e}\left(\points, p' \right)}{\partial t' \, \partial x'_\e} \\
                           \zeros \\
                           \end{array}\right].
\end{multline}

@@

Our motivation for estimating transient strain rates is, in part, to
detect transient geophysical processes. As we will see, geophysical
signal can be easily identified by visually inspecting the solution
for $\strain$ from eqs. (\ref{eq:StrainMean}) and
(\ref{eq:StrainCov}). However, if we want to detect geophysical signal
automatically, then we need to define a detection criterion. We use a
signal-to-noise ratio, SNR, that is based on the Frobenius norm of
$\strain$, $||\strain||_\mathrm{F} = \left(\strain_{\e\e}^2 +
\strain_{\n\n}^2 + 2\strain_{\e\n}^2\right)^{\frac{1}{2}}$, for our
detection criterion. In the geodetic literature,
$||\dot{\varepsilon}||_\mathrm{F}$ is often used as a metric for the
strain rate ``magnitude", and it is sometimes referred to as the
second invariant of strain rate. Noting that $||\strain||_\mathrm{F}$
is a random variable, we take SNR to be the ratio of the estimated
mean and standard deviation of $||\strain||_\mathrm{F}$. An estimate
of the mean is found by evaluating $||\strain||_\mathrm{F}$ at the
mean of $\strain$,
\begin{align}
\mu_{||\strain||_\mathrm{F}} &\approx ||\strain||_F \big|_{\strain = \mu_{\strain}}\nonumber \\
                             &= \left(\mu_{\strain_{\e\e}}^2 + 
                                      \mu_{\strain_{\n\n}}^2 + 
                                      2\mu_{\strain_{\e\n}}^2 \right)^\frac{1}{2},
\end{align}
and we use nonlinear uncertainty propagation to estimate the standard
deviation,
\begin{equation}
\sigma_{||\strain||_\mathrm{F}} \approx
\left( \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\e\e}}\right|_{\strain = \mu_{\strain}} \right)^2 
       \sigma^2_{\strain_{\e\e}} +
       \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\n\n}}\right|_{\strain = \mu_{\strain}} \right)^2
       \sigma^2_{\strain_{\n\n}} +
       \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\e\n}}\right|_{\strain = \mu_{\strain}} \right)^2 
       \sigma^2_{\strain_{\e\n}} \right)^\frac{1}{2},
\end{equation}
where $\sigma^2_{\strain_{ij}}(p) = C_{\strain_{ij}}(p,p)$. After some
calculations, we find SNR to be
\begin{align}\label{eq:SNR}
\mathrm{SNR}(p) &= \frac{\mu_{||\strain||_\mathrm{F}}(p)}{\sigma_{||\strain||_\mathrm{F}}(p)} \\
                &= \frac{\mu_{\strain_{\e\e}}(p)^2 +
                         \mu_{\strain_{\n\n}}(p)^2 +
                         2\mu_{\strain_{\e\n}}(p)^2}
                        {\big(\sigma^2_{\strain_{\e\e}}(p)\mu_{\strain_{\e\e}}(p)^2 + 
                              \sigma^2_{\strain_{\n\n}}(p)\mu_{\strain_{\n\n}}(p)^2 + 
                              4\sigma^2_{\strain_{\e\n}}(p)\mu_{\strain_{\e\n}}(p)^2
                         \big)^{\frac{1}{2}}}.
\end{align}
We explicitly show that SNR is a function of $p$ to emphasize that it
identifies the position and time of anomalous deformation. We can
reasonably suspect that some transient geophysical phenomena is
occurring wherever and whenever SNR is larger than ${\sim}3$.

@@

In this section we estimate transient strain rates in the Pacific
Northwest, and we are specifically interested in identifying transient
strain resulting from SSEs. SSEs in the Pacific Northwest were first
discovered by \citep[e.g.,][]{Dragert2001} and we refer the reader to
\citet{Schwartz2007} for a review of the subsequent research. Before
estimating transient strain rates, we establish a noise model for GNSS
stations in this region, and we establish a prior Gaussian process to
describe displacements from SSEs. SSEs in the Pacific Northwest can be
detected by monitoring for associated seismic tremor
\citep{Rogers2003}, which is actively being done by the Pacific
Northwest Seismic Network \citep{Wech2010}. We can then compare the
tremor records to our estimated transient strain rates to verify that
we are indeed identifying strain from SSEs.

@@

The distribution of inferred values for $\alpha$ and $\beta$ are shown
in Figure \ref{fig:NoiseParams}. The estimates of $\beta$ for the
easting and northing components are clustered around 0.5
mm/yr$^{0.5}$. The corresponding estimates of $\alpha$ tend to cluster
around 0 yr$^{-1}$, indicating that the power spectrum of noise indeed
decays proportionally to $f^{-2}$ over a wide band of frequencies. We
also estimate $\alpha$ and $\beta$ for the vertical component of
displacements, with the hope that vertical deformation gradients could
reveal some geophysical signal. For the vertical component, $\beta$ is
relatively large with a median value of 13.5 mm/yr$^{0.5}$. The
inferred values for $\alpha$ are also higher for the vertical
component with a median value of 8.21 yr$^{-1}$. In Figure
\ref{fig:NoiseSamples}, we use the median values of $\alpha$ and
$\beta$ to generate two realizations of FOGM noise for each component.
The realizations span five years, and the easting and northing
realizations drift by about 1 mm over these five years. In the context
of detecting SSEs, which produce several mm's of surface displacement
on the time-scale of weeks, the estimated FOGM noise for the easting
and northing component is negligible. In contrast, the estimated FOGM
noise for the vertical component is larger than the signal we would
expect from SSEs. We suspect that the higher amplitude for the FOGM
noise in the vertical component is accommodating for deficiencies in
our rather simple seasonal model. Based on this analysis, we
henceforth ignore temporally correlated noise in the easting and
northing component because of its low amplitude. We also do not use
vertical displacements because of the presumably low signal-to-noise
ratio.

@@

Note that $T_i$ includes the parameter $\phi$, which serves to scale
the covariance function $C_{u_i}$. The second option is a member of
the Wendland class of covariance functions \citep{Wendland2005}.
Wendland covariance functions have compact support and hence their
corresponding covariance matrices are sparse. We exploit this sparsity
with the CHOLMOD software package \citep{Chen2008}. Wendland
covariance functions can be constructed such that they are positive
definite in $\mathbb{R}^d$ and their corresponding Gaussian process
can be differentiated $k$ times. We use $d=1$, because we are
describing the temporal covariance of $u_i$, and we use $k=2$, giving
samples of $u_i$ continuous velocities and accelerations. The
corresponding Wendland covariance function is

>>>>

Crustal strain rates are fundamentally important quantities for
assessing seismic hazard. Knowing where and how quickly strain is
accumulating gives insight into where we can expect stored elastic
energy to be released seismically. Estimates of secular, or long-term,
strain rates have been used to constrain seismic hazard models such as
UCERF3 \citep{Field2014}. Transient strain, which is anomalous strain
caused by geophysical processes such as slow slip events (SSEs) or
postseismic deformation, is also relevant for assessing seismic
hazard. While transient strain itself is not damaging, there is a
possibility that it can trigger major earthquakes
\citep{Roeloffs2006,Freed2001}.

@@

The function $C_{u_i}$ must be positive definite in order to be a
valid covariance function. By definition, $C_{u_i}$ is a positive
definite function if the matrix $[C_{u_i}(p,p')]_{(p,p') \in \points
\times \points}$ is positive definite for any set of points $\points$
\citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be
separated into spatial and temporal functions as
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}(p,p') = C_{u_i}\left((\pos,t),(\pos',t')\right) 
              = X_i(\pos,\pos')T_i(t,t').
\end{equation}   
As long as the functions $X_i$ and $T_i$ are positive definite,
$C_{u_i}$ is guaranteed to also be positive definite \citep[sec.
4.2.4]{Rasmussen2006}. We also require that the derivatives
$\partial^2 X_i(\pos,\pos')/ \partial x_j \partial x_j'$ and
$\partial^2 T_i(t,t') / \partial t \partial t'$ exist. This ensures
that $u_i$ is spatially and temporally differentiable, allowing us to
compute transient strain rates (See \citet[sec. 2.2]{Adler1981} or
\citet[appx. 10A]{Papoulis1991} for a definition of stochastic
differentiation and the conditions for differentiability).

@@

We provide an example to give the prior on transient displacements a
more tangible meaning. We can use a squared exponential function for
$X_i$ and $T_i$,
\begin{equation}\label{eq:ExampleCov}
X_i(\pos,\pos') = \exp\left(\frac{-||\pos - \pos'||_2^2}{2\ell^2}\right), \ \ \ \ 
T_i(t,t') = \phi^2 \exp\left(\frac{-|t - t'|^2}{2\tau^2}\right),
\end{equation}
which satisfies our requirements for positive definiteness and
differentiability. The parameters $\ell$ and $\tau$ control the
length-scale and time-scale, respectively, of realizations of $u_i$.
The parameter $\phi$, which we have arbitrarily chosen to incorporate
into $T_i$ rather than $X_i$, controls the amplitude of realizations
of $u_i$. Ideally, we want realizations of $u_i$ to have a
length-scale, time-scale, and amplitude that resemble what we expect
for the true transient displacements. In Figure \ref{fig:PriorDemo}A,
we show $C_{u_i}$ using the squared exponential function for $X_i$ and
$T_i$ and the parameters $\ell = 100$ km, $\tau = 10$ days, and $\phi
= 1$ mm. A single realization of $u_i$ corresponding to that choice of
covariance function is shown in Figure \ref{fig:PriorDemo}B (See
\citet[appx. A.2]{Rasmussen2006} for details on drawing realization from
Gaussian processes). While the squared exponential function is a
commonly used covariance function for GPR, it is not appropriate for
every application. The appropriate choice for $X_i$ and $T_i$ may vary
depending on the geophysical signal we are trying to describe. To keep
this section sufficiently general, we hold off on specifying $X_i$ and
$T_i$ until Section \ref{sec:SignalModel}, where we estimate transient
strain from slow slip events in the Pacific Northwest.

@@

It should be noted that we have ignored any covariances between the
easting and northing components of $\vec{u}$ and $\vec{\data}$. This
simplification reduces the computational complexity of evaluating the
posterior transient displacements because each component can be
evaluated independently. However, we are inherently assuming that the
principal components of $\vec{u}(p)$ and $\vec{d}^{(jk)}$ are aligned
with the cardinal directions, which is admittedly an arbitrary
assumption.

@@

The posterior transient displacements are spatially and temporally
continuous, and we can use eqs. (\ref{eq:PosteriorMean3}) and
(\ref{eq:PosteriorCov3}) to evaluate $\post_i$ at any $p$.
Furthermore, $\post_i$ is spatially and temporally differentiable,
allowing us to formulate $\strain$ at any $p$ that we may be
interested in. The components of $\strain$ can be written as
\begin{equation}\label{eq:StrainRate}
\strain_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} 
                  \left(\frac{\partial \post_i(p)}{\partial x_j} +  
                        \frac{\partial \post_j(p)}{\partial x_i} \right).
\end{equation}
Since eq. (\ref{eq:StrainRate}) is a linear operation on the Gaussian
processes $\post_i$ and $\post_j$, we know that $\strain_{ij}$ is also
a Gaussian process. From \citet[ch. 10]{Papoulis1991}, we find that
the mean and covariance functions for $\dot{\varepsilon}_{ij}$ are
\begin{align}\label{eq:StrainMean}
\mu_{\strain_{\e\e}}(p) &= \frac{\partial^2 \mu_{\post_\e}(p)}{\partial t \, \partial x_\e} \\
\mu_{\strain_{\n\n}}(p) &= \frac{\partial^2 \mu_{\post_\n}(p)}{\partial t \, \partial x_\n} \\
\mu_{\strain_{\e\n}}(p) &= \frac{1}{2}\frac{\partial}{\partial t}\left(
                                    \frac{\partial \mu_{\post_\e}(p)}{\partial x_\n} + 
                                    \frac{\partial \mu_{\post_\n}(p)}{\partial x_\e} \right)
\end{align} 
and
\begin{align}\label{eq:StrainCov}
C_{\strain_{\e\e}}(p,p') &= \frac{\partial^4 C_{\post_\e}(p,p')}{\partial t \, \partial t' \, \partial x_\e \, \partial x'_\e} \\
C_{\strain_{\n\n}}(p,p') &= \frac{\partial^4 C_{\post_\n}(p,p')}{\partial t \, \partial t' \, \partial x_\n \, \partial x'_\n} \\
C_{\strain_{\e\n}}(p,p') &= \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                        \frac{\partial^2 C_{\post_\e}(p,p')}{\partial x_\n \, \partial x'_\n} + 
                                        \frac{\partial^2 C_{\post_\n}(p,p')}{\partial x_\e \, \partial x'_\e} \right),
\end{align} 
respectively. The above derivatives of $\mu_{\post_i}$ and
$C_{\post_i}$ are computed analytically by replacing the terms
$C_{u_i}(p,\points)$, $C_{u_i}(\points,p')$, and $C_{u_i}(p,p')$ in
eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) with their
appropriate derivatives. For example, the mean and covariance
functions for $\strain_{\e\e}$ can be written more verbosely as
\begin{equation}\label{eq:StrainMeanEE}
\mu_{\strain_{\e\e}}(p) = \left[\begin{array}{cc}
                          \frac{\partial^2 C_{u_\e}\left( p,\points \right)}{\partial t \, \partial x_\e} & \zeros \\
                          \end{array}\right]
                          \left[\begin{array}{cc}
                          C_{u_\e}(\points,\points) + \mitbf{C}_{\eta_\e} & \G \\
                          \G^T  & \zeros \\
                          \end{array}\right]^{-1}
                          \left[\begin{array}{c}
                          \data^*_\e \\
                          \zeros \\
                          \end{array}\right]
\end{equation}    
and
\begin{multline}\label{eq:StrainCovEE}
C_{\strain_{\e\e}}(p,p') = \frac{\partial^4 C_{u_\e}(p,p')}{\partial t \, \partial t' \, \partial x_\e \, \partial x'_\e} - \\
                           \left[\begin{array}{cc}
                           \frac{\partial^2 C_{u_\e}\left( p,\points \right)}{\partial t \, \partial x_\e} & \zeros \\
                           \end{array}\right]
                           \left[\begin{array}{cc}
                           C_{u_\e}(\points,\points) + \mitbf{C}_{\eta_\e} & \G \\
                           \G^T  & \zeros \\
                           \end{array}\right]^{-1}
                           \left[\begin{array}{c}
                           \frac{\partial^2 C_{u_\e}\left(\points, p' \right)}{\partial t' \, \partial x'_\e} \\
                           \zeros \\
                           \end{array}\right].
\end{multline}

@@

Our motivation for estimating transient strain rates is, in part, to
detect transient geophysical processes. As we will see, geophysical
signal can be easily identified by visually inspecting the solution
for $\strain$. However, if we want to detect geophysical signal
automatically, then we need to define a detection criterion. We use a
signal-to-noise ratio, SNR, that is based on the Frobenius norm of
$\strain$, $||\strain||_\mathrm{F} = \left(\strain_{\e\e}^2 +
\strain_{\n\n}^2 + 2\strain_{\e\n}^2\right)^{\frac{1}{2}}$, for our
detection criterion. In the geodetic literature,
$||\dot{\varepsilon}||_\mathrm{F}$ is often used as a metric for the
strain rate ``magnitude", and it is sometimes referred to as the
second invariant of strain rate. Noting that $||\strain||_\mathrm{F}$
is a random variable, we take SNR to be the ratio of the estimated
mean and standard deviation of $||\strain||_\mathrm{F}$. An estimate
of the mean is found by evaluating $||\strain||_\mathrm{F}$ at the
mean of $\strain$,
\begin{align}
\mu_{||\strain||_\mathrm{F}} &\approx ||\strain||_F \big|_{\strain = \mu_{\strain}} \nonumber \\
                             &= \left(\mu_{\strain_{\e\e}}^2 + 
                                      \mu_{\strain_{\n\n}}^2 + 
                                      2\mu_{\strain_{\e\n}}^2 \right)^\frac{1}{2},
\end{align}
and we use nonlinear uncertainty propagation to estimate the standard
deviation,
\begin{equation}
\sigma_{||\strain||_\mathrm{F}} \approx
\left( \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\e\e}}\right|_{\strain = \mu_{\strain}} \right)^2 
       \sigma^2_{\strain_{\e\e}} +
       \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\n\n}}\right|_{\strain = \mu_{\strain}} \right)^2
       \sigma^2_{\strain_{\n\n}} +
       \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\e\n}}\right|_{\strain = \mu_{\strain}} \right)^2 
       \sigma^2_{\strain_{\e\n}} \right)^\frac{1}{2},
\end{equation}
where $\sigma^2_{\strain_{ij}}(p) = C_{\strain_{ij}}(p,p)$. After some
calculations, we find SNR to be
\begin{align}\label{eq:SNR}
\mathrm{SNR}(p) &= \frac{\mu_{||\strain||_\mathrm{F}}(p)}{\sigma_{||\strain||_\mathrm{F}}(p)} \nonumber \\
                &= \frac{\mu_{\strain_{\e\e}}(p)^2 +
                         \mu_{\strain_{\n\n}}(p)^2 +
                         2\mu_{\strain_{\e\n}}(p)^2}
                        {\big(\sigma^2_{\strain_{\e\e}}(p)\mu_{\strain_{\e\e}}(p)^2 + 
                              \sigma^2_{\strain_{\n\n}}(p)\mu_{\strain_{\n\n}}(p)^2 + 
                              4\sigma^2_{\strain_{\e\n}}(p)\mu_{\strain_{\e\n}}(p)^2
                         \big)^{\frac{1}{2}}}.
\end{align}
We explicitly show that SNR is a function of $p$ to emphasize that it
identifies the position and time of anomalous deformation. We can
reasonably suspect that some transient geophysical phenomena is
occurring wherever and whenever SNR is larger than ${\sim}3$.

@@

In this section we estimate transient strain rates in the Pacific
Northwest, and we are specifically interested in identifying transient
strain resulting from SSEs. SSEs in the Pacific Northwest were first
discovered by \citet{Dragert2001} and we refer the reader to
\citet{Schwartz2007} for a review of the subsequent research. SSEs in
the Pacific Northwest can be detected by monitoring for associated
seismic tremor \citep{Rogers2003}, which is actively being done by the
Pacific Northwest Seismic Network \citep{Wech2010}. We can then
compare the tremor records to our estimated transient strain rates to
verify that we are indeed identifying strain from SSEs. Before
estimating transient strain rates, we first establish a noise model
for GNSS stations in this region, and then we establish a prior
Gaussian process to describe displacements from SSEs.

@@

The distribution of inferred values for $\alpha$ and $\beta$ are shown
in Figure \ref{fig:NoiseParams}. The estimates of $\beta$ for the
easting and northing components are clustered around 0.5
mm/yr$^{0.5}$. The corresponding estimates of $\alpha$ tend to cluster
around 0 yr$^{-1}$, indicating that the power spectrum of noise indeed
decays proportionally to $f^{-2}$ over a wide band of frequencies. We
also estimate $\alpha$ and $\beta$ for the vertical component of
displacements, with the hope that vertical deformation gradients could
reveal some geophysical signal. For the vertical component, $\beta$ is
relatively large with a median value of 13.5 mm/yr$^{0.5}$. The
inferred values for $\alpha$ are also higher for the vertical
component with a median value of 8.21 yr$^{-1}$. In Figure
\ref{fig:NoiseSamples}, we use the median values of $\alpha$ and
$\beta$ to generate two realizations of FOGM noise for each component.
The realizations span five years, and the easting and northing
realizations drift by about 1 mm over these five years. In the context
of detecting SSEs, which produce several mm's of surface displacement
on the time-scale of weeks, the estimated FOGM noise for the easting
and northing component is negligible. In contrast, the estimated FOGM
noise for the vertical component is larger than the signal we would
expect from SSEs. We suspect that the higher amplitude for the FOGM
noise in the vertical component is accommodating for deficiencies in
our rather simple seasonal model. Based on this analysis, we
henceforth ignore temporally correlated noise in the easting and
northing components because of its low amplitude. We also do not use
vertical displacements because of the presumably low signal-to-noise
ratio.

@@

Note that $T_i$ includes the parameter $\phi$, which serves to scale
the covariance function $C_{u_i}$. A second option is a member of
the Wendland class of covariance functions \citep{Wendland2005}.
Wendland covariance functions have compact support, and thus their
corresponding covariance matrices are sparse. We exploit this sparsity
with the CHOLMOD software package \citep{Chen2008}. Wendland
covariance functions can be constructed such that they are positive
definite in $\mathbb{R}^d$ and their corresponding Gaussian process
can be differentiated $k$ times. We use $d=1$, because we are
describing the temporal covariance of $u_i$, and we use $k=2$, giving
samples of $u_i$ continuous velocities and accelerations. The
corresponding Wendland covariance function is

\begin{equation}\label{eq:Wendland} 
T_i(t,t') = \phi^2\left(1 - \frac{|t - t'|}{\tau}\right)^5_+ 
            \left(\frac{8|t - t'|^2}{\tau^2} + \frac{5|t - t'|}{\tau} + 1\right), 
\end{equation} 
where $(t)_+ = \mathrm{max}(0,t)$. A third option for $T_i$ is the
covariance function for integrated Brownian motion (IBM). IBM is a
Gaussian process with zero mean and its covariance function can be
found by integrating the covariance function for Brownian motion,
\begin{align}\label{eq:IBM}
T_i(t,t') &= \int_0^t \int_0^{t'} \phi^2 \min(s,s') \,ds'\,ds \nonumber \\
          &= \frac{\phi^2}{2}\min(t,t')^2 \left(\max(t,t') - \frac{1}{3}\min(t,t')\right), \ \ \ t,t' \geq 0.
\end{align}
IBM has been used in the context of Kalman filtering as a
non-parametric model for the time dependence of geophysical signals
\citep[e.g.,][]{Segall1997, McGuire2003, Ohtani2010, Hines2016a}.
Similar to Brownian motion, IBM has a reference time, $t=0$, at which
the process begins. For some geophysical signals, it is appropriate to
have this reference time. For example, if we are trying to identify
postseismic deformation, then $t$ should be zero at the time of the
earthquake.  However, if we are interested in detecting transient
events, where there is no known start time, then IBM may not be an
appropriate prior model. Instead, one may prefer to use the SE or
Wendland covariance functions because they are stationary. In the
following analysis, we make the choice that $t$ is zero on the first
epoch of $\data_i^*$. Using an earlier reference time does not change
the results discussed in this section.
